---
title: "Basic Data Analytics"
author: "Cameron Kline-Sharpe"
output: github_document
---

```{r options, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, prompt=TRUE,comment=NULL,message=FALSE, include=TRUE, fig.height = 5, fig.width = 9)
```


```{r getPackages, include=FALSE}
library(tidyverse)
library(CarletonStats)
```

# If you modify this code, please add your name to the top of the document, in the author section.

## TODO:

1. regroup asian, native american with Other for most things
2. For the sake of comparison with our algorithms, focus on making confusion matrix-like graphs
3. Colors often backwards--pick consistant colors; note also to provide colors with correct conotations (red==bad, blue==not)

```{r getData, include=FALSE}
# here we load in the COMPAS data set
# R's automatic loading feature gets the type of some of the data wrong, so those must be set manually 
COMPAS <- read_csv("../CleanedFeaturesData.csv", 
                   col_type = cols(person_id = col_integer(),
                                   age = col_integer(),
                                   juv_fel_count = col_integer(),
                                   juv_misd_count = col_integer(),
                                   juv_other_count = col_integer(),
                                   decile_score = col_integer(),
                                   priors_count = col_integer(),
                                   num_r_cases = col_integer(),
                                   num_vr_cases = col_integer(),
                                   is_recid = col_logical(),
                                   is_violent_recid = col_logical()))

# much of the data in CleanedFeaturesData.csv is not needed, so for the sake of speed
#    we use fewer features of the data
simple_COMPAS <- COMPAS %>% 
  select(person_id, age, race, decile_score, is_recid, is_violent_recid) %>% # get only these columns
  mutate(recidDist = ifelse(is_recid, 10-decile_score, decile_score-1),
         predict5 = ifelse(decile_score > 5, TRUE, FALSE)) %>%  # add a prediction column that predicts recidivism if the score exceeds 5
  filter(recidDist >= 0) %>% # remove all people who have an impossible distance
  mutate(recid = ifelse(is_recid, "Did Recidivate", "Did Not Recidivate"),
         predict5Corr = ifelse(predict5 == is_recid, "Correct", "Incorrect")) # add some strings to the data for ease of display


# read in the Decision tree data for recidivism
DT <- read_csv("DTPredictions.csv",
               col_type = cols(prediction = col_logical(), truth = col_logical())) %>%
      mutate(correct = ifelse(prediction == truth, "Correct", "Incorrect"),
             truthChar = ifelse(truth, "Did Recidivate", "Did Not Recidivate"),
             confusion = ifelse(truth == 1,
                                ifelse(prediction == 1, "TP", "FN"),
                                ifelse(prediction == 0, "TN", "FP")))

# read in the Decision tree data for violent recidivism
DT_Violent <- read_csv("DTViolentPredictions.csv",
               col_type = cols(prediction = col_logical(), truth = col_logical())) %>% # get the data types right
      mutate(correct = ifelse(prediction == truth, "Correct", "Incorrect"),
             truthChar = ifelse(truth, "Did Violently Recidivate", "Did Not Violently Recidivate")) # again, some strings for ease of display


baseline <- read_csv("randomBaselinePredictions.csv",
                     col_type = cols(predicted = col_logical(), truth = col_logical())) %>%
      mutate(correct = ifelse(predicted == truth, "Correct", "Incorrect"),
             truthChar = ifelse(truth, "Did Recidivate", "Did Not Recidivate"))

baseline_violent <- read_csv("randomBaselineResults.csv",
                     col_type = cols(predicted = col_logical(), truth = col_logical())) %>%
      mutate(correct = ifelse(predicted == truth, "Correct", "Incorrect"),
             truthChar = ifelse(truth, "Did Recidivate", "Did Not Recidivate"))

foolish <- read_csv("foolishConditionPredictions.csv",
                     col_type = cols(predicted = col_logical(), truth = col_logical())) %>%
      mutate(correct = ifelse(predicted == truth, "Correct", "Incorrect"),
             truthChar = ifelse(truth, "Did Recidivate", "Did Not Recidivate"))

foolish_violent <- read_csv("foolishConditionResults.csv",
                     col_type = cols(predicted = col_logical(), truth = col_logical())) %>%
      mutate(correct = ifelse(predicted == truth, "Correct", "Incorrect"),
             truthChar = ifelse(truth, "Did Recidivate", "Did Not Recidivate"))

ANN <- read_csv("ANNPred.csv",
                     col_type = cols(prediction = col_logical(), truth = col_logical())) %>%
      mutate(correct = ifelse(prediction == truth, "Correct", "Incorrect"),
             truthChar = ifelse(truth, "Did Recidivate", "Did Not Recidivate"))


green <- "#5BB84C"
green2 <- "#4CB88F"
blue <- "#1FBFC3" 
red <- "#F57670"
redAlarm <- "#E82F16"
```


```{r compasError, echo=FALSE}

simple_COMPAS <- simple_COMPAS %>%
  mutate(pred5Status = ifelse(predict5, 
                              ifelse(is_recid, "True Positive", "False Positive"), 
                              ifelse(is_recid, "False Negative", "True Negative"))) %>% # get the TP, NP, etc.
  mutate(recidStatus = ifelse(is_recid, ifelse(is_violent_recid, "Violently Recidivated", "Did Recidivate"), "Did Not Recidivate")) %>%
  mutate(race3 = ifelse(race %in% c("Asian", "Hispanic", "Native American", "Asian"), "Other", race))

# TODO: try pie chart (maybe just of those who recidivism)
ggplot(simple_COMPAS, aes(x = recidStatus, fill=recidStatus)) + 
  geom_bar() + 
  labs(title = "Reicidivism in our Training Data", x="Did they Recidivate?", fill="Did they Violently Recidivate?") + 
  scale_fill_manual(values = c(green, blue, red))


ggplot(simple_COMPAS, aes(x = recid, fill=recidStatus)) + 
  geom_bar() + 
  labs(title = "Reicidivism in our Training Data (Grouped)", x="Did they Recidivate?", fill="Did they Violently Recidivate?") + 
  scale_fill_manual(values = c(green, blue, red))

ggplot(simple_COMPAS) + 
  geom_bar(mapping = aes(x = recidDist), fill="darkBlue") + 
  labs(title="COMPAS Score Distances", x="COMPAS Score Distances")
  
  
ggplot(simple_COMPAS, aes(x = recidDist, fill=recidStatus)) + geom_bar() +
  facet_wrap(~recidStatus) + 
  labs(title="COMPAS Score Distances by Recidivism", x = "Score Distance", fill = "Recidivism") + 
  scale_fill_manual(values = c(green, blue, red))

simple_COMPAS %>%
  ggplot(aes(x = recidDist, fill=recidStatus)) + geom_bar() +
    facet_wrap(~race3) + 
    labs(title="COMPAS Score Distances by Race", x = "Score Distance", fill = "Recidivism") + 
    scale_fill_manual(values = c(green, blue, red))

simple_COMPAS %>%
  ggplot(aes(x = decile_score, fill=recidStatus)) + geom_bar() +
    facet_wrap(~race3) + 
    labs(title="COMPAS Scores by Race", x = "Score Distance", fill = "Recidivism") + 
    scale_fill_manual(values = c(green, blue, red))

simple_COMPAS %>%
  filter(race == "African-American") %>%
  ggplot(aes(x = decile_score, fill=recidStatus)) + geom_bar() +
    labs(title="COMPAS Scores for African Americans", x = "Score Distance", fill = "Recidivism") + 
    scale_fill_manual(values = c(green, blue, red))

simple_COMPAS %>%
  filter(race == "Caucasian") %>%
  ggplot(aes(x = decile_score, fill=recidStatus)) + geom_bar() +
    labs(title="COMPAS Scores for Caucasians", x = "Score Distance", fill = "Recidivism") + 
    scale_fill_manual(values = c(green, blue, red))

simple_COMPAS %>%
  filter(race == "Hispanic") %>%
  ggplot(aes(x = decile_score, fill=recidStatus)) + geom_bar() +
    labs(title="COMPAS Scores for Hispanic People", x = "Score Distance", fill = "Recidivism") + 
    scale_fill_manual(values = c(green, blue, red))

simple_COMPAS %>%
  filter(race %in% c("Other", "Asian", "Native American")) %>%
  ggplot(aes(x = decile_score, fill=recidStatus)) + geom_bar() +
    labs(title="COMPAS Scores for Other Racial Groups", x = "Score Distance", fill = "Recidivism") + 
    scale_fill_manual(values = c(green, blue, red))

# TODO: check that these are good colors
ggplot(simple_COMPAS, aes(x = pred5Status, fill=pred5Status)) +
  geom_bar() + 
  theme(legend.position = "none") + 
  labs(title = "Confusion Matrix as a Bar Graph", x="Prediction Status") + 
  scale_fill_manual(values = c(red, redAlarm, green, green2))
```

### CONFUSION MATRIX
  Confusion Matrix | Score > 5 | Score <= 5 |
------------------ | --------- | ---------- |
Did Recidivate     |   5333    | 1993       |
Did not Recidivate |   1794    | 1909       |


```{r race, echo = FALSE}


simple_COMPAS <- simple_COMPAS %>%
  mutate(race2 = ifelse(race %in% c("Asian", "Native American"), "Other", race))

ggplot(simple_COMPAS, aes(x=race2, fill=is_recid)) + 
  geom_bar() + 
  labs(title = "Proportion of Recidivism by Race (Grouped) [Compare with Previous]", x="Grouped Race", y="Number", fill="Recidivism")

ggplot(simple_COMPAS, aes(x=race, fill=recidStatus)) + 
  geom_bar() + 
  labs(title = "Proportion of Recidivism by Race", x="Race", y="Number", fill="Recidivism") + 
  scale_fill_manual(values = c(green, blue, red)) 

ggplot(simple_COMPAS, aes(x=race2, fill=is_recid)) + 
  geom_bar() + 
  labs(title = "Proportion of Recidivism by Race (Grouped)", x="Grouped Race", y="Number", fill="Recidivism")

ggplot(simple_COMPAS, aes(x=race2, fill=recidStatus)) + 
  geom_bar() + 
  labs(title = "Proportion of Recidivism by Race (Grouped) [Compare with Previous]", x="Grouped Race", y="Number", fill="Recidivism") + 
  scale_fill_manual(values = c(green, blue, red)) 

ggplot(simple_COMPAS, aes(x=race2, y=decile_score, color = race2)) + 
  geom_boxplot() + 
  labs(title = "COMPAS Decile Score by Race", y="Decile Score", x="Race (Grouped)", color="Race") +
  stat_summary(fun.y = mean, geom="point", shape = 22)

ggplot(simple_COMPAS, aes(x=race2, color=race2, y=recidDist)) + 
  geom_boxplot() + 
  labs(title="COMPAS Score Distance by Race", y="Distance", x="Race (Grouped)") + 
  stat_summary(fun.y = mean, geom="point", shape = 22)

simple_COMPAS %>%
  ggplot(mapping = aes(x=race, y=recidDist, color=recid)) + 
    geom_boxplot() + 
    labs(title="Compas Score Distance by Race and Recidivism", y="Distance", x="Race (Grouped)")
```

```{r age, echo=FALSE}
by_age <- simple_COMPAS %>% 
  group_by(age) %>%
  summarise(numRecid=sum(is_recid), numViolent=sum(is_violent_recid), n=n()) %>%
  gather(2:4, key = metric, value = n)

ggplot(by_age, aes(x = age, y = n, color = metric)) + 
  geom_path() + 
  labs(title = "Number of People by Age", x = "Age", y="Count")


simple_COMPAS %>%
  group_by(age) %>%
  summarize(mean_score = mean(decile_score)) %>%
  ggplot(aes(x=age, y=mean_score)) + geom_path() +
    labs(title = "Mean COMPAS score by Age")

ggplot(simple_COMPAS, aes(x=age, y=recidDist)) + 
  geom_smooth() +
  labs(title = "COMPAS Score Distance by Age", y = "COMPAS Score Distance")

ggplot(simple_COMPAS, aes(x=age, y=recidDist, color=recidStatus)) + 
  geom_smooth() +
  labs(title = "COMPAS Score Distance by Age and Recidivism", y = "COMPAS Score Distance") + 
  scale_color_manual(values = c(green, blue, red)) 

```


```{r setup, echo=FALSE}

simple_COMPAS <- simple_COMPAS %>%
  mutate(recid = ifelse(is_recid, "Did Recidivate", "Did Not Recidivate"),
         correct = ifelse(predict5 == is_recid, "Correct", "Incorrect"),
         truth = is_recid)

# scale things to 1.0

names <- c("Random")
accs <- c(0.5)
TPRs <- c(0.5)
TNRs <- c(0.5)
FPRs <- c(0.5)

vnames <- c("Random")
vaccs <- c(0.5)
vTPRs <- c(0.5)
vTNRs <- c(0.5)

accuracy <- function(data){
  nrow(filter(data, correct == "Correct"))/nrow(data)
}

TPR <- function(data){
  nrow(filter(data, correct == "Correct", truth == 1))/nrow(filter(data, truth == 1))
}

TNR <- function(data){
  nrow(filter(data, correct == "Correct", truth == 0))/nrow(filter(data, truth == 0))
}

FPR <- function(data){
  nrow(filter(data, truth == 0, correct == "Incorrect"))/nrow(filter(data, truth == 0))
}


con_mat <- function(data){
  data %>%
    group_by(truthChar) %>%
    mutate(num = n()) %>%
    group_by(truthChar, correct, num) %>%
    summarize(n = n()) %>%
    mutate(prop = round(n/num,3))
}

get_fairness <- function(old, model, whiteFPR, blackFPR){
  new <- list(Model = model, White = whiteFPR, Black = blackFPR) %>%
    as.data.frame() %>%
    gather(White, Black, key=Race, value = FPR)
  data <- rbind(old, new)
  data
}

plot_fairness <- function(data){
  ggplot(data, aes(x=Model, y=FPR, fill = Race)) +
    geom_bar(stat = "identity", color = "black", position="dodge") +
    scale_fill_manual(values = c(blue, green)) +
    ylim(0,1)
}


plot_metrics <- function(names, acc, TPR, FPR){
  met <- list(Model=names, acc=acc, TPR=TPR, FPR=FPR) %>%
         as.data.frame() %>%
         gather(acc, TPR, FPR, key=Metric, value=Proportion)
  ggplot(met, aes(x=Model, y=Proportion, fill=Metric)) +
    geom_bar(stat="identity", position="dodge",
             color="black") +  
    ylim(0,1)
}

plot_accs <- function(names, accs){
  met <- list(Model = names, Accuracy = accs) %>%
    as.data.frame() %>%
    as.tbl()
  ggplot(met, aes(x=Model, y=Accuracy, fill = Model)) +
    geom_bar(stat="identity", color = "black") +
    geom_text(aes(label = round(Accuracy, 3), y = Accuracy + 0.05),
              alpha = 0.9, size = 6) +
    ylim(0,1)
}

```


```{r baseline}


simple_COMPAS %>%
  ggplot( aes(x=recid, fill=predict5Corr)) + 
    geom_bar() + 
    scale_fill_manual(values = c("#1FBFC3", "#F57670")) + 
    labs(title="COMPAS Confusion Matrix", fill="COMPAS Prediction")

names <- append(names, "COMPAS")
accs <- append(accs, accuracy(simple_COMPAS))
FPRs <- append(FPRs, FPR(simple_COMPAS))

baseline %>%
  con_mat() %>%
  ggplot(aes(x = truthChar, y=n, fill=correct)) + 
    geom_bar(stat="identity") +
    labs(title = "Random Baseline for Recidivism",
         fill="Prediction",
         x = "Ground Truth",
         y = "Count") + 
    scale_fill_manual(values = c("#1FBFC3", "#F57670")) + 
    geom_text(aes(label = prop, y=n), size = 6, position = position_stack(vjust = 0.5))

              
baseline_violent %>%
  con_mat() %>%
  ggplot(aes(x = truthChar, y=n, fill=correct)) + 
    geom_bar(stat="identity") +
    labs(title = "Random Baseline for Violent Recidivism",
         fill="Prediction",
         x = "Ground Truth",
         y = "Count") + 
    scale_fill_manual(values = c("#1FBFC3", "#F57670")) + 
    geom_text(aes(label = prop, y=n), size = 4, position = position_stack(vjust = 0.5))

plot_accs(names, accs) + 
  scale_fill_manual(values = c(green, blue)) 

foolish %>%
  con_mat() %>%
  ggplot(aes(x = truthChar, y=n, fill=correct)) + 
    geom_bar(stat="identity") +
    labs(title = "Foolish Prediction for Recidivism",
         fill="Prediction",
         x = "Ground Truth",
         y = "Count") + 
    scale_fill_manual(values = c("#1FBFC3", "#F57670")) + 
    geom_text(aes(label = prop, y=n), size = 6, position = position_stack(vjust = 0.5))

names <- append(names, "Foolish")
accs <- append(accs, accuracy(foolish))
FPRs <- append(FPRs, FPR(foolish))
plot_accs(names, accs) + 
  scale_fill_manual(values = c(green, blue, blue)) 


foolish_violent %>%
  con_mat() %>%
  ggplot(aes(x = truthChar, y=n, fill=correct)) + 
    geom_bar(stat="identity") +
   
    labs(title = "Foolish Prediction for Violent Recidivism",
         fill="Prediction",
         x = "Ground Truth",
         y = "Count") + 
    scale_fill_manual(values = c("#1FBFC3", "#F57670")) + 
    geom_text(aes(label = prop, y=n), size = 4, position = position_stack(vjust = 0.5))

vnames <- append(vnames, "Foolish")
vaccs <- append(vaccs, accuracy(foolish_violent))
vTPRs <- append(vTPRs, TPR(foolish_violent))
vTNRs <- append(vTNRs, TNR(foolish_violent))



```


```{r algorithms}

NB <- simple_COMPAS %>%
  mutate(truth = is_recid) %>%
  select(truth) %>%
  mutate(NB = "Did Not Recidivate",
         truthChar= ifelse(truth == 1, "Did Recidivate", "Did Not Recidivate"),
         correct = ifelse(truthChar == NB, "Correct", "Incorrect"))

NB %>%
  con_mat() %>%
  ggplot(aes(x = truthChar, y=n, fill=correct)) + 
    geom_bar(stat="identity") +
    labs(title = "Naive Bayes Prediction for Recidivism",
         fill="Prediction",
         x = "Ground Truth",
         y = "Count") + 
    scale_fill_manual(values = c("#1FBFC3", "#F57670")) + 
    geom_text(aes(label = prop, y=ifelse(n < 500, 500, n)),
              size = 6, position = position_stack(vjust = 0.5))

names <- append(names, "NB")
accs <- append(accs, accuracy(NB))
FPRs <- append(FPRs, FPR(NB))
plot_accs(names, accs) +
  scale_fill_manual(values = c(green, blue, blue, blue))


DT %>%
  con_mat() %>%
  ggplot(aes(x = truthChar, y=n, fill=correct)) + 
    geom_bar(stat="identity") +
    labs(title = "Decision Tree Prediction for Recidivism",
         fill="Prediction",
         x = "Ground Truth",
         y = "Count") + 
    scale_fill_manual(values = c("#1FBFC3", "#F57670")) + 
    geom_text(aes(label = prop, y=ifelse(n < 500, 500, n)),
              size = 6, position = position_stack(vjust = 0.5))

names <- append(names, "DT")
accs <- append(accs, accuracy(DT))
FPRs <- append(FPRs, FPR(DT))
plot_accs(names, accs) +
  scale_fill_manual(values = c(blue,green, blue, blue, blue))


DT_Violent %>%
  con_mat() %>%
  ggplot(aes(x = truthChar, y=n, fill=correct)) + 
    geom_bar(stat="identity") +
    labs(title = "Decision Prediction for Violent Recidivism",
         fill="Prediction",
         x = "Ground Truth",
         y = "Count") + 
    scale_fill_manual(values = c("#1FBFC3", "#F57670")) + 
    geom_text(aes(label = prop, y=ifelse(n < 500, 500, n)),
              size = 6, position = position_stack(vjust = 0.5))

vnames <- append(vnames, "DT")
vaccs <- append(vaccs, accuracy(DT_Violent))
vTPRs <- append(vTPRs, TPR(DT_Violent))
vTNRs <- append(vTNRs, TNR(DT_Violent))

ANN %>%
  con_mat() %>%
  ggplot(aes(x = truthChar, y=n, fill=correct)) + 
    geom_bar(stat="identity") +
    labs(title = "ANN Prediction for Recidivism",
         fill="Prediction",
         x = "Ground Truth",
         y = "Count") + 
    scale_fill_manual(values = c("#1FBFC3", "#F57670")) + 
    geom_text(aes(label = prop, y=ifelse(n < 500, 500, n)),
              size = 6, position = position_stack(vjust = 0.5))

names <- append(names, "ANN")
accs <- append(accs, accuracy(ANN))
FPRs <- append(FPRs, FPR(ANN))
plot_accs(names, accs) + 
  scale_fill_manual(values = c(green, blue, blue, blue, blue, blue))

  
```
```{r fairness}


plot_accs(names, FPRs) +
  labs(x = "False Positive Rate")

Fair_by_race <- list(Model = c("Random", "Random"),
             Race = c("White", "Black"),
             FPR = c(0.5, 0.5)) %>%
  as.data.frame() %>%
  as.tbl()


Fair_by_race <- get_fairness(Fair_by_race, "COMPAS",
                             FPR(filter(simple_COMPAS, race == "Caucasian")),
                             FPR(filter(simple_COMPAS, race == "African-American")))

Fair_by_race <- get_fairness(Fair_by_race, "DT",
             FPR(filter(DT, race == "Caucasian")),
             FPR(filter(DT, race == "African-American")))

Fair_by_race <- get_fairness(Fair_by_race, "ANN",
                             FPR(filter(ANN, Caucasian == 1)),
                             FPR(filter(ANN, `African-American` == 1)))


plot_fairness(Fair_by_race)

Fair_by_race %>%
  spread(Race, FPR) %>%
  mutate(FPR_Difference = Black - White) %>%
  ggplot(aes(x = Model, y = FPR_Difference)) +
    geom_bar(stat = "identity", color = "black",
             position="dodge", fill = red) +
    geom_text(aes(label = round(FPR_Difference, 3),
                  y = FPR_Difference + 0.008),
              size = 6)

Sex_fair <- list(Model = c("Random", "Random"),
             Race = c("White", "Black"),
             FPR = c(0.5, 0.5)) %>%
  as.data.frame() %>%
  as.tbl()

Sex_fair <-  get_fairness(Sex_fair, "ANN",
             FPR(filter(ANN, Male == 1)),
             FPR(filter(ANN, Female == 1)))


```

```{r boots, eval=FALSE}

# African American 

b_vs_w <- simple_COMPAS %>%
  filter(race %in%  c("Caucasian", "African-American")) %>%
  select(race,decile_score)

b_vs_everyone <- simple_COMPAS %>%
  mutate(not_black = ifelse(race == "African-American", "AA", "Other")) %>%
  select(not_black, decile_score)

boot(decile_score ~ race, b_vs_w) # 95 % CI = 1.563831 1.782578 (AA _ C)

boot(decile_score ~ not_black, b_vs_everyone) #  95 % CI =  1.728979 1.933784 (AA - E)

# Hispanic

h_vs_w <- simple_COMPAS %>%
  filter(race %in%  c("Caucasian", "Hispanic")) %>%
  select(race,decile_score)

h_vs_everyone <- simple_COMPAS %>%
  mutate(not_black = ifelse(race == "Hispanic", "H", "Other")) %>%
  select(not_black, decile_score)


boot(decile_score ~ race, h_vs_w) # 95 % CI = 0.1633147 0.5171069 (C - H)

boot(decile_score ~ not_black, h_vs_everyone) #  95 % CI =  -1.377856 -1.036370 (H - E)

# Age ranges

age_groups <- simple_COMPAS %>%
  mutate(ageYoung = ifelse(age <= 35, "Y", "O"), ageOld = ifelse(age >= 55, "O", "Y")) %>%
  select(ageYoung, ageOld, decile_score)

boot(decile_score ~ ageYoung, age_groups) #  95 % CI =  -2.061035 -1.854572 (O - Y)

boot(decile_score ~ ageOld, age_groups) #  95 % CI =  -2.152014 -1.815392 (O - Y)


```























